---
title: "NYC_Predictions"
author: "Gosuddin Siddiqi"
date: "February 19, 2017"
output: pdf_document
---

```{r setup, warning=FALSE,message=FALSE}
library(dplyr)
library(geosphere)
library(lubridate)
library(rpart)
library(randomForest)

#load data
nyc_sept_data <- read.csv("green_tripdata_2015-09.csv")


```


#Analysing Data

```{r}
#retreive column names
colnames(nyc_sept_data)

#summary
summary(nyc_sept_data)

#remove Ehail
nyc_sept_data$Ehail_fee <- NULL

#remove nas
dataset <- na.omit(nyc_sept_data)

```

#Cleaning Data

```{r}
#remove negative values # Impossible to exists like negative distance, fares
dataset$Fare_amount <-
ifelse(dataset$Fare_amount<0,-(dataset$Fare_amount),dataset$Fare_amount)

dataset$Extra <- ifelse(dataset$Extra<0,-(dataset$Extra),dataset$Extra)

dataset$MTA_tax <- ifelse(dataset$MTA_tax<0,-(dataset$MTA_tax),dataset$MTA_tax)

dataset$Tip_amount <- ifelse(dataset$Tip_amount<0,-(dataset$Tip_amount),dataset$Tip_amount)

dataset$Tolls_amount <- ifelse(dataset$Tolls_amount<0,-(dataset$Tolls_amount),dataset$Tolls_amount)

dataset$improvement_surcharge <- ifelse(dataset$improvement_surcharge<0,-(dataset$improvement_surcharge),dataset$improvement_surcharge)

dataset$Total_amount <- ifelse(dataset$Total_amount<0,-(dataset$Total_amount),dataset$Total_amount)

#summary(dataset)

hist(dataset$Passenger_count)

hist(dataset$Fare_amount,breaks = 10)
hist(log(dataset$Fare_amount))

boxplot(log(dataset$Fare_amount), outline = F)


#count records with no GPS coordinates (0,0)
n<-dataset%>%
  filter(Pickup_longitude == 0 | Pickup_latitude == 0 |
           Dropoff_longitude == 0 | Dropoff_latitude == 0)%>%
  summarise(count_0 = n())

#remove the records where there is no change in coordinates and trip_distance is 0
z_dist_index <- which(dataset$Trip_distance ==0)

for( i in 1:length(z_dist_index)){
  ind <- z_dist_index[i]
   x1 <- dataset$Pickup_latitude[ind]
   y1 <- dataset$Pickup_longitude[ind]
   x2 <- dataset$Dropoff_latitude[ind]
   y2 <- dataset$Dropoff_longitude[ind]
  dataset$Trip_distance[ind] <- distHaversine(c(y1,x1),c(y2,x2)) / 1609.34
}

#find indexes
rem <- (which(dataset$Trip_distance ==0))

#remove 
dataset <- dataset[-rem,]

#remove records with no GPS coordinates (0,0)
rem <- which((dataset$Pickup_longitude == 0 | dataset$Pickup_latitude == 0 |
                  dataset$Dropoff_longitude == 0 | dataset$Dropoff_latitude == 0))

dataset <- dataset[-rem,]

#remove rows for which amount does not exists
rem<-which(dataset$Tip_amount == 0 & dataset$Total_amount == 0)

dataset <- dataset[-rem,]

#Convert date from factors to POSIX
dataset$day_of_month_p <- mday( dataset$lpep_pickup_datetime)
dataset$day_of_week_p <- wday( dataset$lpep_pickup_datetime)
dataset$day_of_month_d <- mday( dataset$Lpep_dropoff_datetime)
dataset$day_of_week_d <- wday( dataset$Lpep_dropoff_datetime)

dataset$hour_p <- hour(dataset$lpep_pickup_datetime)
dataset$hour_d <- hour(dataset$Lpep_dropoff_datetime)

#create a feature tip as percentage

dataset<-dataset %>%
  mutate(tip_percentage = ((Tip_amount/Total_amount)*100))


#Whether the driver recieved tip?
dataset <- dataset %>%
  mutate(gave_tip = ifelse(tip_percentage == 0,1,0))

dataset <-dataset %>%
            mutate(time_elapsed_mins = difftime(Lpep_dropoff_datetime,lpep_pickup_datetime, units ="mins"))

#Launch the speed variable
dataset$time_elapsed_mins <- as.numeric(dataset$time_elapsed_mins)


```

```{r}


with_zero_time <-which(dataset$time_elapsed_mins == 0)

summary(dataset[with_zero_time,])



```
Seems like Verifone isn't good at reporting the data. All of the data which shows 0 mins for the ride were recorded by the Verifone.

```{r}
#Removing the data that was inserted possibly randomly

dataset <- dataset[-which((dataset$time_elapsed_mins == 0) & hour(dataset$lpep_pickup_datetime)==0 & hour(dataset$Lpep_dropoff_datetime) == 0),]

dataset <- dataset[-which(dataset$time_elapsed_mins ==0),]

```


```{r}
#Would NYC taxis really operate beyond 50 miles?
summary(dataset[which(dataset$Trip_distance>50),])

#seems like there are few which actually goes beyond 50 miles. Maybe we should just discard it.

dataset <- dataset[-which(dataset$Trip_distance>50),]

dataset <- dataset %>%
  mutate(avg_speed = Trip_distance/(time_elapsed_mins/60))


summary(dataset)
```

Since we know that there is no information about the cash tips, it is very unlikely that the passengers who paid the total fare in cash would have paid the tip amount in different mode of payment. So my model would be based on the assumption that if I am to predict an entry for which I know the mode of payment was cash, I would neglect it.
```{r}
which(dataset$Payment_type ==2 & dataset$Tip_amount != 0)

#let's check rows with cash payment yet tip amount above 0
dataset[-which(dataset$Payment_type ==2 & dataset$Tip_amount != 0),]

#removing them
dataset <- dataset[-which(dataset$Payment_type ==2 & dataset$Tip_amount != 0),]
```
It seems like there are only just 2 cases where we have mode of payment as cash and still some tip was awarded to the driver.


Removing the entries for which we cannot predict the tip amount as the payment type is cash (payment type == 2)
```{r}
dataset <- dataset[-which(dataset$Payment_type ==2),]


#tips only from credit card
dataset <- dataset[which(dataset$Payment_type ==1),]

```



```{r}
summary(dataset[which(dataset$avg_speed>80),])


dataset <- dataset[-which(dataset$avg_speed>80),]

dataset <-dataset[-which(dataset$tip_percentage == 100),]

summary(dataset$Payment_type)
```

```{r}
names(dataset)

unique(dataset$Payment_type)


dataset$lpep_pickup_datetime <- as.POSIXct(dataset$lpep_pickup_datetime)
dataset$Lpep_dropoff_datetime <- as.POSIXct(dataset$Lpep_dropoff_datetime)
dataset$RateCodeID<-as.factor(dataset$RateCodeID)
dataset$Trip_type <- as.factor(dataset$Trip_type)
dataset$Payment_type <- as.factor(dataset$Payment_type)
dataset$day_of_week_d <- as.factor(dataset$day_of_week_d)
dataset$day_of_week_p <- as.factor(dataset$day_of_week_p)
dataset$hour_p <- as.factor(dataset$hour_p)
dataset$hour_d <- as.factor(dataset$hour_d)
```
```{r}
indep <- c("RateCodeID","Pickup_longitude","Pickup_latitude","Dropoff_longitude","Dropoff_latitude","Passenger_count","Trip_distance","Fare_amount","Extra","MTA_tax","Tolls_amount","improvement_surcharge","Payment_type","Trip_type","day_of_month_p","day_of_week_p",
           "hour_p","time_elapsed_mins")

dep <- c("gave_tip")

sample <- sample(1:nrow(dataset), 0.85*nrow(dataset)) 

train <-(dataset[sample,])
test <-(dataset[-sample,])

str(dataset)

dc_fit <- rpart (gave_tip ~
                Dropoff_latitude+
                Dropoff_longitude+
                Pickup_latitude+
                Pickup_longitude+
                avg_speed+
                time_elapsed_mins+
                Trip_distance+
                Lpep_dropoff_datetime+
                lpep_pickup_datetime+
                hour_d+
                hour_p
                
              ,data=train, method = "class" )


summary(dc_fit)


predicted= predict(dc_fit,test, type = "class")

length(which(test$gave_tip == predicted))
```

```{r}
mod <- lm(tip_percentage ~ Dropoff_latitude+
                Dropoff_longitude+
                Pickup_latitude+
                Pickup_longitude+
                avg_speed+
                time_elapsed_mins+
                Trip_distance+
                Lpep_dropoff_datetime+
                lpep_pickup_datetime+
                hour_d+
                hour_p, data = train )

summary(mod)

plot(mod$fitted.values)
```

```{r}
#random forest attempt

fit <- randomForest(gave_tip ~.
              ,data=train[which(train$RateCodeID != 5),!names(train) %in% c("tip_percentage","Tip_amount","Total_amount")], ntree =100)

summary(fit)

predicted <- predict(fit,test)

length(which(test$gave_tip == predicted))

#sqrt(mean((test$tip_percentage - predicted)^2))

rr<-roc(fit$predicted,train$gave_tip )
plot(rr)

#mod <- lm(tip_percentage ~ Trip_distance + time_elapsed_mins + Fare_amount, data = train)

#summary(mod)

#predicted <-predict(mod,test)

#sqrt(mean((test$tip_percentage - predicted)^2))
```



#Extracting important features
```{r}
#remove categorical index
cat_index<- c("lpep_pickup_datetime","Lpep_dropoff_datetime","Store_and_fwd_flag","VendorID", "RateCodeID","gave_tip","Payment_type","Trip_type")

#compute PCA
pca_res <- prcomp(dataset[-which(dataset$time_elapsed_mins==0),!names(dataset) %in% cat_index])


```

```{r}

dataset <- dataset[-which(dataset$Payment_type ==2 | dataset$time_elapsed_mins>240),]



mod <- glm(as.factor(gave_tip)~ Dropoff_latitude+
                Dropoff_longitude+
                Pickup_latitude+
                Pickup_longitude+
                avg_speed+
                time_elapsed_mins+
                Trip_distance+
                Lpep_dropoff_datetime+
                hour_p +
                day_of_week_p
              ,family = "binomial", data=train[,!colnames(train)%in%c("tip_percentage","Tip_amount",
                                                "Total_amount")])

summary(mod)

fits <- fitted(mod)

y <- factor(train$gave_tip)


library(AUC)
rr <- roc(fits,y)

plot(rr)

auc(rr)

sample_index <- sample(1:nrow())
test <- 

res<-predict(mod,test[,!colnames(test)%in%"gave_tip"],type='response')

plot(log(dataset$tip_percentage))

hist(dataset$time_elapsed_mins)

hist(log(dataset$tip_percentage),breaks = 20)

plot(dataset$Tolls_amount, dataset$gave_tip)

summary(dataset[-which(dataset$Payment_type == 2),])
```

